(working title) Local Physiological Adaptation in \textit{Mimulus cardinalis}

### 2015-04-27 ms/ms.Rnw
Worked on ms

@@

As with previous studies in \textit{Mimulus} and many other species [CITES], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g. temperature or precipitation) that covary with latitude.

MARKER - continue here!!!

Methods specific to latitude vs. trait correlation:
We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We used a comprehensive set of \cdm{358 in total before thinning} recent (since 2000) known \textit{M. cardinalis} occurences recorded from herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included X environmental variables from ClimateWNA and 17? bioclimatic variables (Table S\#). For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation). Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.

>>>>

<<thinnedOccClimate>>=
# R data.frame of 178 M. cardinalis occurrences, thinned to correct for sampling bias, and associated climate data
load(paste(pathDat, "/thinnedOccClimate.RData", sep = ""))
@

@@

As with previous studies in \textit{Mimulus} and many other species [CITES], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g. temperature or precipitation) that covary with latitude. We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We did this by looking for overlap between climatic variables that best predict latitude of \textit{M. cadinalis} populations and climatic variables that best predict trait variation across our 16 focal populations. We used a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M. cardinalis} occurences. These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included X environmental variables from ClimateWNA and 17? bioclimatic variables (Table~\ref{table:TableS_}). For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation). Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.

### 2015-04-27 ms/ms.tex
Worked on ms

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
  \hlcom{# The path to the ms directory may need to be changed for different users}
  \hlcom{# Manuscript files}
  \hlstd{pathMS} \hlkwb{<-} \hlstr{"~/Google Drive/CardLocalAdaptation/ms"}

@@

We used 2 watering treatments on two populations on two soils. The watering treatments are: well-watered, drought.

\subsubsection*{Temperature}

We simulated typical growing season (June and July) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (High Temp) and Little Jameson (Low Temp). We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}. Daily temperatures from ClimateWNA are usually highly correlated with the air temperature recorded from data loggers in the field at these sites (A. Angert, unpub. data). Hence, the ClimateWNA temperature profiles are likely to similar to actual thermal regimes experienced by \textit{M. cardinalis} in nature. To create realistic temperature regimes, we calculated the mean temperature trend from June to July using LOESS \citep{Cleveland_etal_1992}. The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ). The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals. We fit and simulated from the VAR(2) using the package \pkg{dse} \citep{Gilbert_2014} in R. Simulated data closely resembled the autocorrelation and between-site correlation of the actual data. From simulated mean temperature, we next selected minimum and maximum daily temperatures. Mean, min, and max temperature were highly correlated at both sites. We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models. For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature. During the middle of the day, chamber temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.

\subsubsection*{Water}

      April 14 germinate
      Before May 20 (36 days): 7.5 min (complete flood) every 2 hours
      between May 20 to June 10 (57 days), down to 30 seconds, 1x per day
			Wet: daily/constant irrigation with cooled water
			Dry: start at week \#, gradually decrease watering frequency/level

@@

\cdm{which traits: germination, growth rate, biomass, biomass allometry, leaf traits, photosynthesis?, mortality?}
We tested for Population, Treatment, and Population $\times$ Treatment interactions using mixed model ANOVA with Family included as a random factor. Models were fit by restricted maximum likelihood using lmer from the R package \pkg{lme4} \citep{Bates_etal_2014}. Significant fixed effect terms were selected using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-11 \citep{Kuznetsova_etal_2014}. Denominator degrees of freedom for $F$-tests were estimated using Satterthwaite's approximation. Significant Population effect indicate intrinsic trait differences; significant Population $\times$ Treatment effects indicate population differences in plasticity. For growth rate, we also accounted for differences in germination rate by including day of germination as a factor.

\subsection*{Principal components of germination, growth, and photosynthesis}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
  \hlstd{popMeans} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}
    \hlkwc{germ} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{1}\hlstd{],} \hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{1}\hlstd{]} \hlopt{+} \hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{2}\hlopt{:}\hlnum{16}\hlstd{]),}
    \hlkwc{LER} \hlstd{= fitLER}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{],}
    \hlkwc{SER} \hlstd{= fitSER}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{],}
    \hlkwc{photo} \hlstd{= fitIntrnPhoto}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{])}

@@

@@

As with previous studies in \textit{Mimulus} and many other species [CITES], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g. temperature or precipitation) that covary with latitude.

@@

Methods specific to latitude vs. trait correlation:
We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We used a comprehensive set of \cdm{358 in total before thinning} recent (since 2000) known \textit{M. cardinalis} occurences recorded from herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included X environmental variables from ClimateWNA and 17? bioclimatic variables (Table S\#). For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation). Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.

>>>>

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
  \hlcom{# The path to the ms directory may need to be changed for different users}
  \hlcom{# Manuscript files}
  \hlstd{pathMS} \hlkwb{<-} \hlstr{"~/Google Drive/CardLocalAdaptation/ms"} \hlcom{# Chris' MacBook}
  \hlcom{# pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop}

@@

We imposed four treatments, a fully-factorial cross of two temperature treatments and two watering treatments. Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment and another to the Cool treatment. Within each chamber, there were two Well-watered blocks and two Drought blocks. A detailed description of treatments is given below \cdm{should some be moved to supplement?} and summarized in Fig~\ref{fig:Fig_ExptlDes}. The irradiance in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$. The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature treatments (data not shown).

\subsubsection*{Temperature treatments}

We simulated typical growing season (June 1 - August 15) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (Hot) and Little Jameson (Cool). We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}. Daily temperatures from ClimateWNA are highly correlated with the air temperature recorded from data loggers in the field at these sites (A. Angert, unpub. data). Hence, the ClimateWNA temperature profiles are similar to actual thermal regimes experienced by \textit{M. cardinalis} in nature. We simulated realistic temperature regimes by calculating the mean temperature trend from June to August using LOESS \citep{Cleveland_etal_1992}. The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ). The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals. We fit and simulated from the VAR(2) model using the package \pkg{dse} \citep{Gilbert_2014} in R. Simulated data closely resembled the autocorrelation and between-site correlation of the actual data. From simulated mean temperature, we next selected minimum and maximum daily temperatures. Mean, min, and max temperature were highly correlated at both sites. We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models. For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature. During the middle of the day, temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.

\subsubsection*{Watering treatments}

For watering treatments, we simulated two extreme types of streams where \textit{M. cardinalis} grows. In the well-watered treatment, we simulated a large stream that never goes dry during the summer growing season. In the drought treatment, we simulated a small stream that has ample flow at the beginning of the season, but gradually dries down as the winter snow pack melts. In both treatments, plants were bottom-watered using  water chilled to 7.5 [check] by MAKE AND MODEL OF CHILLER. Plants in the well-watered treatment were fully saturated every two hours during the day. Watering in the drought treatment gradually declined from every two hours to every day between May 20 (36 days after sowing) and 10 June (57 days after sowing). Simultaneously, the amount of bottom-watering per flood decreased, such that only the bottom of the cone-tainers were wetted by the end of the experiment.

@@

For all traits (Table~\ref{table:Table_traits}) we tested for Population, Treatment, and Population $\times$ Treatment interactions. We interpreted significant Population effects to indicate intrinsic variation and Population by Treatment effects to indicate variation in plasticity. As mentioned above, we survival and GLM models for germination rate and mortality, respectively. For all other traits, we used mixed model ANOVAs with Family included as a random factor. Models were fit by restricted maximum likelihood using lmer from the R package \pkg{lme4} \citep{Bates_etal_2014}. Significant fixed effect terms were selected using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-11 \citep{Kuznetsova_etal_2014}. Denominator degrees of freedom for $F$-tests were estimated using Satterthwaite's approximation. Significant Population effect indicate intrinsic trait differences; significant Population $\times$ Treatment effects indicate population differences in plasticity. For growth rate, we also accounted for differences in germination rate by including day of germination as a factor.

\subsection*{Principal components of germination, growth, and photosynthesis}
[maybe this goes in Results?]
We summarized population-level coefficients, after factoring out Treatment and other effects, using principal components. Spec
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
  \hlstd{popMeans} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}
    \hlkwc{germ} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{1}\hlstd{],} \hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{1}\hlstd{]} \hlopt{+} \hlkwd{coef}\hlstd{(fitGerm)[}\hlnum{2}\hlopt{:}\hlnum{16}\hlstd{]),}
    \hlkwc{LER} \hlstd{= fitLER}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{],}
    \hlkwc{SER} \hlstd{= fitSER}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{],}
    \hlkwc{photo} \hlstd{= fitIntrnPhoto}\hlopt{$}\hlstd{lsmeans.table[}\hlnum{1}\hlopt{:}\hlnum{16}\hlstd{,} \hlstr{"Estimate"}\hlstd{])}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# R data.frame of 178 M. cardinalis occurrences, thinned to correct for sampling bias, and associated climate data}
\hlkwd{load}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/thinnedOccClimate.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

As with previous studies in \textit{Mimulus} and many other species [CITES], latitude \textit{per se} predicted trait variation more than many obvious environmental variables (e.g. temperature or precipitation) that covary with latitude. We used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We did this by looking for overlap between climatic variables that best predict latitude of \textit{M. cadinalis} populations and climatic variables that best predict trait variation across our 16 focal populations. We used a representative set of 178 recent (since 2000) known \textit{M. cardinalis} occurences. These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included X environmental variables from ClimateWNA and 17? bioclimatic variables (Table~\ref{table:TableS_}). For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation). Temperatures were converted to Kelvin to be on an interval scale appropriate for calculating the coefficient of variation.

@@
